---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggnetwork

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![GitHub tag](https://img.shields.io/github/tag/briatte/ggnetwork.svg?label=latest)](https://github.com/briatte/ggnetwork/releases)
[![R-CMD-check](https://github.com/briatte/ggnetwork/workflows/R-CMD-check/badge.svg)](https://github.com/briatte/ggnetwork/actions?workflow=R-CMD-check)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/briatte/ggnetwork?branch=master&svg=true)](https://ci.appveyor.com/project/briatte/ggnetwork)
[![Coverage Status (codecov)](https://codecov.io/gh/briatte/ggnetwork/branch/master/graph/badge.svg)](https://codecov.io/gh/briatte/ggnetwork)
[![CRAN status](https://www.r-pkg.org/badges/version-ago/ggnetwork)](https://cran.r-project.org/package=ggnetwork)
[![CRAN checks (worst)](https://cranchecks.info/badges/worst/ggnetwork)](https://cran.r-project.org/web/checks/check_results_ggnetwork.html)
[![CRAN total downloads](https://cranlogs.r-pkg.org/badges/grand-total/ggnetwork)](https://cran.r-project.org/package=ggnetwork)
<!-- badges: end -->

This package allows to pass network objects to [`ggplot2`](https://ggplot2.org/) and provides geometries to plot their elements.

## Get started

You can install the released version of `ggnetwork` [from CRAN](https://cran.r-project.org/package=ggnetwork) with:

``` r
install.packages("ggnetwork")
```

And the development version [from GitHub](https://github.com/briatte/ggnetwork) with:

``` r
# install.packages("remotes")
remotes::install_github("briatte/ggnetwork")
```

The `ggnetwork` package depends on __R 3.5+__ and on `ggplot2` version 2.0.0+.

## Documentation

- The [package vignette](https://briatte.github.io/ggnetwork/) contains detailed examples of how to use its `fortify` method and each of its geometries.
- The [package website](https://briatte.github.io/ggnetwork/articles/ggnetwork.html) has more information and a list of [all package functions](https://briatte.github.io/ggnetwork/reference/index.html).

For further examples that use `ggnetwork` with other packages to produce animated graphs, see James Curley's slides on “[Interactive and Dynamic Network Visualization in R](http://curleylab.psych.columbia.edu/netviz/)” (2016). For even more options, see Katherine Ognyanova's tutorial “[Network visualization with R](https://kateto.net/network-visualization) (2019), and David Schoch's guide “[Network Visualizations in R using `ggraph` and `graphlayouts`](http://mr.schochastics.net/netVizR.html)” (2019).

## Getting help

If you encounter a clear bug, please file a minimal reproducible example [on GitHub](https://github.com/briatte/ggnetwork/issues).

For questions and other discussion, please contact the package maintainer.

## Citation

You can get a citation for the package from R:

``` r
citation("ggnetwork")
```

## See also

The `ggnetwork` package was written within a larger development effort around network visualization with `ggplot2`, on which you can read the following article:

> Sam Tyner, François Briatte and Heike Hofmann, “[Network Visualization with `ggplot2`](https://doi.org/10.32614/RJ-2017-023),” _The R Journal_ 9(1): 27–59, 2017.

The article also covers the related packages [`geomnet`](https://github.com/sctyner/geomnet) and  [`ggnet`](https://github.com/briatte/ggnet). It does not cover the more recent [`ggraph`](https://github.com/thomasp85/ggraph), [`graphlayouts`](https://github.com/schochastics/graphlayouts) and [`tidygraph`](https://github.com/thomasp85/tidygraph), although you should turn to those if you need a highly extensive way to build and plot 'tidy' networks with `ggplot2`.

## Thanks

```{r, include = FALSE, eval = FALSE}
usethis::use_tidy_thanks()
```

Thanks to 
[&#x0040;achmurzy](https://github.com/achmurzy), [&#x0040;andrewd789](https://github.com/andrewd789), [&#x0040;ArtemSokolov](https://github.com/ArtemSokolov), [&#x0040;aterhorst](https://github.com/aterhorst), [&#x0040;emillykkejensen](https://github.com/emillykkejensen), [&#x0040;evinhas](https://github.com/evinhas), [&#x0040;FinScience](https://github.com/FinScience), [&#x0040;ghost](https://github.com/ghost), [&#x0040;instantkaffee](https://github.com/instantkaffee), [&#x0040;jalapic](https://github.com/jalapic), [&#x0040;jcfisher](https://github.com/jcfisher), [&#x0040;jfaganUK](https://github.com/jfaganUK), [&#x0040;kippjohnson](https://github.com/kippjohnson), [&#x0040;koheiw](https://github.com/koheiw), [&#x0040;mbojan](https://github.com/mbojan), 
[&#x0040;mcanouil](https://github.com/mcanouil), [&#x0040;mgagliol](https://github.com/mgagliol), [&#x0040;mhairi](https://github.com/mhairi), [&#x0040;minimaxir](https://github.com/minimaxir), [&#x0040;mkarikom](https://github.com/mkarikom), [&#x0040;nick-youngblut](https://github.com/nick-youngblut), [&#x0040;SantiFilippo](https://github.com/SantiFilippo), [&#x0040;sciabolazza](https://github.com/sciabolazza), [&#x0040;sctyner](https://github.com/sctyner), [&#x0040;trinker](https://github.com/trinker), [&#x0040;zachcp](https://github.com/zachcp) and two anonymous _[R Journal](https://journal.r-project.org/)_ reviewers.
[&#x0040;heike](https://github.com/heike) and 
[&#x0040;ethen8181](https://github.com/ethen8181) also helped with the tricky issue of having arrows on directed edges, while 
[&#x0040;sumtxt](https://github.com/sumtxt) inspired this package as well as its predecessor, the `ggnet` package, which also benefitted from discussions with 
[&#x0040;pedroj](https://github.com/pedroj) and Bertrand Sudre.

---
Please note that this project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md).  
By participating in this project you agree to abide by its terms.
