---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# ggnetwork

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![GitHub tag](https://img.shields.io/github/tag/briatte/ggnetwork.svg?label="latest tag")](https://github.com/briatte/ggnetwork)
[![Travis-CI Build Status](https://travis-ci.org/briatte/ggnetwork.svg?branch=master)](https://travis-ci.org/briatte/ggnetwork)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/briatte/ggnetwork?branch=master&svg=true)](https://ci.appveyor.com/project/briatte/ggnetwork)
[![Coverage Status (codecov)](https://codecov.io/gh/briatte/ggnetwork/branch/master/graph/badge.svg)](https://codecov.io/gh/briatte/ggnetwork)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version-ago/ggnetwork)](https://cran.r-project.org/package=ggnetwork)
[![cran checks_worst](https://cranchecks.info/badges/worst/ggnetwork)](https://cran.r-project.org/web/checks/check_results_ggnetwork.html)
[![CRAN_Download_total](http://cranlogs.r-pkg.org/badges/grand-total/ggnetwork)](https://cran.r-project.org/package=ggnetwork)
<!-- badges: end -->

This package allows to pass network objects to [`ggplot2`](http://ggplot2.org/), along with several geometries to plot their elements.

## Get started

You can install the released version of ggnetwork from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("ggnetwork")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("briatte/ggnetwork")
```

__IMPORTANT: the `ggnetwork` package depends on `ggplot2` version 2.0.0 or above.__

__The package vignette__ contains [detailed examples](https://briatte.github.io/ggnetwork/) of how to use its fortify method and each of its geometries.

__For further examples__ that use `ggnetwork` with other packages to produce animated graphs, see James Curley's slides on "[Interactive and Dynamic Network Visualization in R](http://curleylab.psych.columbia.edu/netviz/)."

---

## Getting help

If you encounter a clear bug, please file a minimal reproducible example on [github](https://github.com/briatte/ggnetwork/issues).  
For questions and other discussion, please contact the package maintainer.

## See also

Related packages:

* [geomnet](https://github.com/sctyner/geomnet)
* [ggnet](https://github.com/briatte/ggnet)
* [ggraph](https://github.com/thomasp85/ggraph)

## Thanks

```{r, include = FALSE, eval = FALSE}
usethis::use_tidy_thanks()
```

Thanks to two anonymous _[R Journal](https://journal.r-project.org/)_ reviewers,
[&#x0040;achmurzy](https://github.com/achmurzy), [&#x0040;andrewd789](https://github.com/andrewd789), [&#x0040;ArtemSokolov](https://github.com/ArtemSokolov), [&#x0040;aterhorst](https://github.com/aterhorst), [&#x0040;briatte](https://github.com/briatte), [&#x0040;emillykkejensen](https://github.com/emillykkejensen), [&#x0040;evinhas](https://github.com/evinhas), [&#x0040;FinScience](https://github.com/FinScience), [&#x0040;ghost](https://github.com/ghost), [&#x0040;instantkaffee](https://github.com/instantkaffee), [&#x0040;jalapic](https://github.com/jalapic), [&#x0040;jcfisher](https://github.com/jcfisher), [&#x0040;jfaganUK](https://github.com/jfaganUK), [&#x0040;kippjohnson](https://github.com/kippjohnson), [&#x0040;koheiw](https://github.com/koheiw), [&#x0040;mbojan](https://github.com/mbojan), [&#x0040;mcanouil](https://github.com/mcanouil), [&#x0040;mgagliol](https://github.com/mgagliol), [&#x0040;mhairi](https://github.com/mhairi), [&#x0040;minimaxir](https://github.com/minimaxir), [&#x0040;mkarikom](https://github.com/mkarikom), [&#x0040;nick-youngblut](https://github.com/nick-youngblut), [&#x0040;SantiFilippo](https://github.com/SantiFilippo), [&#x0040;sciabolazza](https://github.com/sciabolazza), [&#x0040;sctyner](https://github.com/sctyner), [&#x0040;trinker](https://github.com/trinker) and [&#x0040;zachcp](https://github.com/zachcp)


---
Please note that this project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md).  
By participating in this project you agree to abide by its terms.
