---
title: 'ggnetwork: Network geometries for ggplot2'
author: "François Briatte"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: default
    toc: yes
vignette: >
  %\VignetteIndexEntry{ggnetwork}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, message=FALSE}
library(ggplot2)
library(ggnetwork)
```

> The [`ggnetwork`](https://github.com/briatte/ggnetwork) package provides a way to build network plots with [`ggplot2`](https://ggplot2.tidyverse.org/).

Install the stable version from CRAN:

```{r, eval=FALSE}
install.packages("ggnetwork")
```

Or use `remotes` to install the latest version of the package [from GitHub](https://github.com/briatte/ggnetwork):

```{r, eval=FALSE}
remotes::install_github("briatte/ggnetwork")
```

The package is meant to be used with `ggplot2` version 2.0.0 or above, so make sure that you update your version of `ggplot2` from CRAN before using `ggnetwork`:

```{r, eval=FALSE}
install.packages("ggplot2")
library(ggplot2)
```

`ggnetwork` further requires the `network` and `sna` packages for network manipulation, and will also install the [`ggrepel`](https://github.com/slowkow/ggrepel) package for repulsive label drawing.

The `ggnetwork` package is very much related to the development of [`geom_net`](https://github.com/sctyner/ggnet) by [Samantha C. Tyner](https://github.com/sctyner) and [Heike Hoffmann](https://github.com/heike). It also shares some similarity to the [`ggnet` and `ggnet2` functions](https://github.com/briatte/ggnet), which are part of the [`GGally`](https://github.com/ggobi/ggally) package by Barret Schloerke and others. Each of these projects are extensions to Hadley Wickham's implementation of Leland Wilkinson's "grammar of graphics" in [ggplot2](https://ggplot2.tidyverse.org/).

## Minimal example

Let's define a small random graph to illustrate each component of `ggnetwork`:

```{r, message=FALSE}
library(network)
library(sna)
n <- network(rgraph(10, tprob = 0.2), directed = FALSE)
```

Let's now add categorical and continuous attributes for both edges and vertices. We'll start with nodes, adding a categorical vertex attribute called `"family"`, which is set to either `"a"`, `"b"` or `"c"`, and a continuous vertex attribute called `"importance"`, which is set to either 1, 2 or 3.

```{r}
n %v% "family" <- sample(letters[1:3], 10, replace = TRUE)
n %v% "importance" <- sample(1:3, 10, replace = TRUE)
```

We now add a categorical edge attribute called `"type"`, which is set to either `"x"`, `"y"` or `"z"`, and a continuous vertex attribute called `"day"`, which is set to either 1, 2 or 3.

```{r}
e <- network.edgecount(n)
set.edge.attribute(n, "type", sample(letters[24:26], e, replace = TRUE))
set.edge.attribute(n, "day", sample(1:3, e, replace = TRUE))
```

Last, note that `ggnetwork` contains a "blank" plot theme that will avoid plotting axes on the sides of the network. We will use that theme in most of the plots:

```{r}
theme_blank
```

## Main building blocks

### `ggnetwork`

The `ggnetwork` package is organised around a ‘workhorse’ function of the same name, which will ‘flatten’ the network object to a data frame that contains the edge list of the network, along with the edge attributes and the vertex attributes of the sender nodes.

The network object referred to above might be an object of class `network`, or any data structure that can be coerced to it, such as an edge list, an adjacency matrix or an incidence matrix. If the `intergraph` package is installed, then objects of class `igraph` can also be used with the `ggnetwork` package.

The data frame returned by `ggnetwork` also contains the coordinates needed for node placement as columns `"x"`, `"y"`, `"xend"` and `"yend"`, which as a consequence are "reserved" names in the context of `ggnetwork`. If these names show up in the edge or the vertex attributes, the function will simply fail to work.

The default node placement algorithm used by `ggnetwork` to produce these coordinates is the Fruchterman-Reingold force-directed layout algorithm. All of the [placement algorithms implemented in the `sna` package](https://www.rdocumentation.org/packages/sna/versions/2.5/topics/gplot.layout) are available through `ggnetwork`, which also accepts additional layout parameters:

```{r, eval=FALSE}
ggnetwork(n, layout = "fruchtermanreingold", cell.jitter = 0.75)
ggnetwork(n, layout = "target", niter = 100)
```

The `layout` argument will also accept user-submitted coordinates as a two-column matrix with as many rows as the number of nodes in the network.

The top of the data frame produced by `ggnetwork` contains self-loops to force every node to be included in the plot. This explains why the rows shown below have the same values in `"x"` and `"xend"` (and in `"y"` and `"yend"`), and only missing values in the columns corresponding to the edge attributes:

```{r}
head(ggnetwork(n))
```

The next rows of the data frame contain the actual edges:

```{r}
tail(ggnetwork(n))
```

The data frame returned by `ggnetwork` has (_N_ + _E_) rows, where _N_ is the number of nodes of the network, and _E_ its number of edges. This data format is very likely to include duplicate information about the nodes, which is unavoidable.

Note that `ggnetwork` does not include any safety mechanism against duplicate column names. As a consequence, if there is both a vertex attribute called `"na"` and an edge attribute called `"na"`, as in the example above, then the vertex attribute will be renamed `"na.x"` and the edge attribute will be renamed `"na.y"`.

### `fortify.network` and `fortify.igraph`

The ‘flattening’ process described above is implemented by `ggnetwork` as `fortify` methods that are recognised by `ggplot2`. As a result, `ggplot2` will understand the following syntax as long as `n` is an object of class `network` or of class `igraph`:

```{r, eval=FALSE}
ggplot(n)
```

However, if the object `n` is a matrix or an edge list to be coerced to a network object, you are required to use the `ggnetwork` function to pass the object to `ggplot2`:

```{r, eval=FALSE}
ggplot(ggnetwork(n))
```

### `geom_edges`

Let's now draw the network edges using `geom_edges`, which is just a lightly hacked version of `geom_segment`. In the example below, we map the `type` edge attribute to the linetype of the network edges:

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(aes(linetype = type), color = "grey50") +
  theme_blank()
```

The other aesthetics that we mapped are the basic coordinates of the network plot. These might also be set as part of the call to `geom_segment`, but setting them at the root of the plot avoids having to repeat them in additional geoms.

Note that `geom_edges` can also produce curved edges by setting its `curvature` argument to any value above `0` (the default):

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(aes(linetype = type), color = "grey50", curvature = 0.1) +
  theme_blank()
```

### `geom_nodes`

Let's now draw the nodes using `geom_nodes`, which is just a lightly hacked version of `geom_point`. In the example below, we map the `family` vertex attribute to the color of the nodes, and make the size of these nodes proportional to the `importance` vertx attribute:

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(aes(linetype = type), color = "grey50") +
  geom_nodes(aes(color = family, size = importance)) +
  theme_blank()
```

Because `ggplot2` follows Wilkinson's grammar of graphics, it accepts only one color scale. In the example above, that scale is mapped to a vertex attribute, but it could have also been mapped to an edge attribute. Mapping a color to _both_ a vertex attribute and an edge attribute will create a single color scale that incorrectly merges both attributes into one:

```{r, eval=FALSE}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(aes(color = type)) +
  geom_nodes(aes(color = family)) +
  theme_blank()
```

This is a limitation of `ggnetwork` that would require violating some fundamental aspects of the grammar of graphics to be circumvented.

## More building blocks

### `geom_nodetext`

Let's now add node labels. These are simply plotted over the nodes by the `nodetext` geom, which works exactly like `geom_text`. In the example below, we map the `vertex.names` attribute (which contains numbers 1 to 10) to uppercase letters:

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(color = "black") +
  geom_nodes(color = "black", size = 8) +
  geom_nodetext(aes(color = family, label = LETTERS[ vertex.names ]),
                fontface = "bold") +
  theme_blank()
```

### `geom_nodelabel`

If you prefer to use the `geom_label` geom recently introduced in `ggplot2`, `ggnetwork` also supports these through the `nodelabel` geom:

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(color = "black") +
  geom_nodelabel(aes(color = family, label = LETTERS[ vertex.names ]),
                 fontface = "bold") +
  theme_blank()
```

### `geom_nodetext_repel` and `geom_nodelabel_repel`

`ggnetwork` supports the repulsive label functions introduced by the `ggrepel` package, which allows to label nodes with non-overlapping annotations. Simply add `_repel` to either `geom_nodetext` or `geom_nodelabel` to use that functionality:

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(color = "black") +
  geom_nodelabel_repel(aes(color = family, label = LETTERS[ vertex.names ]),
                       fontface = "bold", box.padding = unit(1, "lines")) +
  geom_nodes(color = "black", size = 8) +
  theme_blank()
```

### `geom_edgetext` and `geom_edgelabel`

Let's now add edge labels. These are plotted at mid-distance of the nodes that the edges connect by the `edgetext` geom, which works exactly like `geom_label`, except that its default arguments do not draw a border around the labels. Here's an example where we map the `day` edge attribute to edge labels:

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(aes(linetype = type), color = "grey75") +
  geom_nodes(color = "gold", size = 8) +
  geom_nodetext(aes(label = LETTERS[ vertex.names ])) +
  geom_edgetext(aes(label = day), color = "white", fill = "grey25") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        panel.background = element_rect(fill = "grey25"),
        panel.grid = element_blank())
```

The `edgelabel` geom is just an alias of the `edgetext` geom. Note that these geoms are unlikely to produce adequate results if the edges produced by `geom_edges` are curved.

### `geom_edgetext_repel` and `geom_edgelabel_repel`

As you would do with nodes, simply add `_repel` to either `geom_edgetext` or `geom_edgelabel` to draw repulsive edge labels:

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(aes(linetype = type), color = "grey75") +
  geom_nodes(color = "gold", size = 8) +
  geom_nodetext(aes(label = LETTERS[ vertex.names ])) +
  geom_edgetext_repel(aes(label = day), color = "white", fill = "grey25",
                      box.padding = unit(1, "lines")) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        panel.background = element_rect(fill = "grey25"),
        panel.grid = element_blank())
```

## More plotting parameters

This section presents some rather experimental features of `ggnetwork`.

### Edge arrows

`ggnetwork` uses code by [Heike Hoffmann](https://github.com/heike) to better show arrows in directed graphs. To illustrate this, we will need a directed graph example, so let's use the first of the seven `emon` networks bundled in the `network` package:

```{r}
data(emon)
emon[[1]]
```

If this network is passed to `ggnetwork` without any further plotting parameter, the result will feature "shortened" edges that do not reach their receiver nodes:

```{r, echo=FALSE}
ggplot(emon[[1]], aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges() +
  geom_nodes(color = "tomato", size = 4) +
  theme_blank()
```

This is because directed networks are expected to be plotted with edge arrows indicating the directedness of each edge. Adding edge arrows with `geom_edges` works through the same call to the `arrow` function that is supported by `geom_segment`:

```{r}
ggplot(emon[[1]], aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(arrow = arrow(length = unit(6, "pt"), type = "closed")) +
  geom_nodes(color = "tomato", size = 4) +
  theme_blank()
```

The slightly shortened edges avoid overplotting the edge arrows and the nodes. The amount of "edge shortening" can be set through the `arrow.gap` parameter of `ggnetwork`, which defaults to `0` when the network is undirected and `0.025` when it is. This parameter might need adjustment depending on the size of the nodes, and it will probably not manage to avoid any overplotting when the size of the nodes is not constant.

### Edge weights

`ggnetwork` can use an edge attribute as edge weights when computing the network layout. The name of that edge attribute should be passed to the `weights` argument for that to happen, as in this example, which will produce different layouts than if `weights` had been left set to `NULL` (the default):

```{r, eval=FALSE}
ggnetwork(emon[[1]], weights = "Frequency")
```

The Kamada-Kawai is one example of a network layout that supports edge weights. The user should refer to the documentation of each network layout to understand which of these can make use of edge weights.

If `ggnetwork` finds duplicated edges in a network, it will return a warning, as these edges should probably have been converted to single weighted edges for adequate plotting.

### Node faceting

In order for `ggnetwork` to operate correctly with faceted plots, the `by` argument, which is `NULL` by default, can be set to the name of an edge attribute. The result will be a longer data frame that can be plotted with either `facet_wrap` or `facet_grid`, as in the example below, where the faceting variable, the `Frequency` edge attribute, has to be specified twice (once to `ggnetwork`, once to `facet_wrap`):

```{r}
ggplot(ggnetwork(emon[[1]], arrow.gap = 0.04, by = "Frequency"),
       aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(arrow = arrow(length = unit(6, "pt"), type = "closed"),
             aes(color = Sponsorship)) +
  geom_nodes(aes(color = Sponsorship), size = 4) +
  facet_wrap(~ Frequency) +
  theme_facet()
```

The `by` argument is basically an attempt to bring minimal support for temporal networks in `ggnetwork`. It will systematically show all nodes in all plot facets, using the same coordinates in every facet. For more advanced plots of dynamic networks, the user should turn to the [`ndtv`](https://cran.r-project.org/package=ndtv) and [`tsna`](https://cran.r-project.org/package=tsna) packages.

The example above also shows how to use a vertex attribute as part of the aesthetics of the edges. Given how `ggnetwork` operates, these vertex attributes will always be those of the sender node.

Last, the example also shows that `ggnetwork` comes with a theme called `theme_facet`. This theme is a variation of the previously mentioned `theme_blank` that preserves its facet boxes:

```{r}
theme_facet
```

## Additional methods

Since `ggnetwork` works entirely through `ggplot2`, all `ggplot2` methods apply:

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(aes(linetype = type), color = "grey50") +
  geom_nodes(aes(x, y, color = family, size = 1.5 * importance)) +
  geom_nodetext(aes(label = LETTERS[ vertex.names ], size = 0.5 * importance)) +
  geom_edgetext(aes(label = day), color = "grey25") +
  scale_color_brewer(palette = "Set2") +
  scale_size_area("importance", breaks = 1:3, max_size = 9) +
  theme_blank()
```

Similarly, it is possible to use any of the geometries more than once per plot:

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(color = "grey50", alpha = 0.5) +
  geom_nodes(aes(x, y, color = family, size = 5.5 * importance), alpha = 0.5) +
  geom_nodes(aes(x, y, color = family, size = 1.5 * importance)) +
  scale_color_brewer(palette = "Set1") +
  guides(size = FALSE) +
  theme_blank()
```

Last, all geoms provided by `ggnetwork` can be subsetted through the `data` argument, just as any `ggplot2` geom, and as in the example below, which draws only a subset of all node labels:

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(color = "grey50", alpha = 0.5) +
  geom_nodes(aes(x, y, color = family), size = 3) +
  geom_nodelabel_repel(aes(label = vertex.names),
                       box.padding = unit(1, "lines"),
                       data = function(x) { x[ x$family == "a", ]}) +
  scale_color_brewer(palette = "Set1") +
  theme_blank()
```

---

```{r, results='asis', echo=FALSE}
cat("Last printed on ", gsub("\\s+", " ", format(Sys.time(), "%b %e, %Y")), 
    ", using ggnetwork version ", as.character(packageVersion("ggnetwork")),
    ".", sep = "")
```
